Master's Project Journal

### 2019-04-23 thesis/FPGABasedADPLLs.tex
partial corrections

\section{Chapter Overview}
The first step in creating \iac{FPGA} based network of \acp{ADPLL} is the design of the \ac{ADPLL} itself, which will be addressed in this chapter. The nature of \iac{FPGA} necessitates a number of compromises in the design of a given block which limits transferability to \ac{ASIC} designs. In this chapter the potential designs for each individual block, or module, investigated will be explained and the case for their selection in \iac{FPGA} based \ac{ADPLL} made. A number of blocks have implement purely digital circuitry and as such can be transferred in their entirety from \iac{FPGA} and vice versa. However, those that will be used to emulate mixed-signal circuitry, such as the \ac{DCO} and \ac{PFD}, will be examined in greater detail. \Iac{RTL} diagram of the \ac{ADPLL} topology that will be assumed in the following sections is shown in Figure \ref{fig:network_adpll}.
\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{../network_adpll}
	\caption{\ac{ADPLL} designed for use in a Cartesian grid network.}
	\label{fig:network_adpll}
\end{figure}

Important to note is that despite a Cartesian grid leading to four neighbours, only two phase detectors are required per oscillator. Borrowing matrix indexing, and choosing the oscillator located at index $i,j$, $i,j \neq 1,N$ where $N$ is the number of oscillators in each direction, comparison need only be done with the signals generated by oscillators located at $i-1,j$ and $i,j-1$. The comparison with oscillators $i+1,j$ and $i,j+1$ will be carried out in those \acp{ADPLL}, and the negation of the errors measured used in the \ac{ADPLL} located at $i,j$.

\section{Digitally Controlled Oscillators}
The choices made in the design of the \ac{DCO} have the greatest impact on the effectiveness of the overall platform and which use cases the \ac{ADPLL} containing it are suitable for, as the key performance benchmarks are all done using the waveform this block generates. This project will address three distinct designs of \acp{ADPLL} suitable for implementation on an FPGA, two derived from the clocks generated by the \acp{FPGA} own distribution network and one generated independently of this clock, using a chain of inverters. These are not the only ways in which an oscillator could be synthesised on \iac{FPGA}, however other designs were deemed to be unsuitable for extensible and portable implementations.

A prime example of this is the use of Xilinx proprietary \texttt{IODELAY} blocks to create an oscillator, as detailed in Xilinx Application Note XAPP872 \cite{iodelay}. The key idea here is that the bulk of the period is made up by the propagation time through one of the \texttt{IODELAY} blocks, which can be set at implementation time. This is combined with a section of an inverter chain, and a multiplexer used modify the length of this segment, the output of which is fed back into the \texttt{IODELAY} block. This method was discarded as the number of \texttt{IODELAY} blocks is very limited, so expanding to a larger network would be impossible, and they are all located around the edge of the chip, not suited to the construction of a Cartesian grid.

The main issue with the creation of \acp{DCO} on \iac{FPGA} is the inability to create mixed-signal circuits, such as those that would be intended for use on \iac{ASIC}. As such the \ac{FPGA} based oscillator must emulate the behaviour of a mixed-signal circuit in some way.

\subsection{\acs{FPGA} Driven, Linear Period \acs{DCO}}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../osc1}
	\caption{\acs{FPGA} driven, linear period \acs{DCO} \ac{RTL} Diagram.}
	\label{fig:osc1}
\end{figure}
The first design of \ac{DCO} to be examined is of the type used by Zianbetov in his \ac{ADPLL} network test bed and relies on a counter driven by the clock manager on the \ac{FPGA} \cite{zianbetov2013phd}. At each event on the \ac{FPGA} provided clock a counter is incremented, overflowing upon increment past the maximum possible value. The \ac{MSB} of this counter forms the waveform generated by this oscillator, the period of which is controlled through an adjustable value that is loaded into the counter once overflow is reached, and forms the starting points for the counter. The period of oscillation is given by:
\begin{equation}
	T_{osc} = \big(2^{width} -~(BIAS+CC)~\big)\times T_{FPGA}
\end{equation}
where $T_{FPGA}$ is the clock period of the \ac{FPGA}, $CC$ is the control code input, $BIAS$ centres the oscillator in the middle of the tuning range in the event that the control code is zero and $width$ is the width of the counter used. As the only variable here is the control code, period step of this design is $T_{FPGA}$, thereby providing period linearity with respect to the control code. This is the key advantage of this design, as most \ac{ASIC} implementations of \iac{DCO} are also linear in period. The other main reason to choose this design is that its \ac{FPGA} clocked nature allows for exact control over the frequency of operation, and the number of tunable parameters make it possible to configure multiple ways to achieve the same frequencies of operation. Combined these attributes make it very easy to create an oscillator that emulates the behaviour of a design intended for \iac{ASIC}, however, at a greatly reduced frequency. This restriction on the frequency of operation arises out of the period step size, which in order to obtain a good resolution must be orders of magnitude smaller than the intended period to be generated. As the output waveform is taken from the counter's \ac{MSB}, the reload value of the counter must never go beyond $2^{width-1}$, as otherwise the output waveform will become a constant 1. As the reload value varies the low time of the \ac{MSB}, if the desired output waveform is a square wave this design will not be suitable.

In being \ac{FPGA} clocked this design has pseudo-deterministic characteristics, with each period step being almost identical across oscillators and over the entire tuning range, unlike \iac{ASIC} where process variation will impact the layout of a high frequency oscillator. The only variation in this design will come, ironically, from jitter or skew in the \ac{FPGA}'s clock distribution network, which as the frequencies will typically be in the low hundreds of MHz is very minor. In the case of the Xilinx \acl*{Nexys} this is at most 100 picoseconds, or 0.05\% of the period of an intended output clock at 5 MHz. To put this value into perspective, on this board the minimum value of $T_{FPGA}$ that could be used to drive this oscillator is 3.87 nanoseconds, 1.935\% of the period.

@@

\subsection{\acs{FPGA} Driven, Linear Frequency \acs{DCO}}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../osc2}
	\caption{\acs{FPGA} driven, linear frequency \acs{DCO} \ac{RTL} Diagram.}
	\label{fig:osc2}
\end{figure}
The second \ac{FPGA} clocked oscillator is similar in most attributes to the above design but eschews period linearity for frequency linearity. Again the overflow property of a counter is used with the counter's \ac{MSB} as output of the block, however, this time it forms a square wave. Rather than setting the reload value of the counter, instead the increment is adjusted depending on the control code, thus requiring $\frac{2^N}{BIAS+CC}$ increments to overflow. Accordingly the frequency of operation is set by:
\begin{equation}
	f_{osc} = f_{FPGA}\times\frac{BIAS+CC}{2^{width}}
\end{equation}
Here the control code $CC$ and bias are added to the value stored in the accumulator at each event of the \ac{FPGA}, clock until overflow is reached. This occurs at $2^{width}$ where, as before, $width$ is the bit width of the counter, thus valuing each control code increment at $\frac{f_{FPGA}}{2^{width}}$ Hz. As with the previous design, this oscillator is better suited to frequencies where the output of the \ac{DCO} is orders of magnitude lower than the clock signal driving it, as this ensures that the incremental change due to the control code remains a small fraction of the period.

This design is just as configurable as its linear-in-period counterpart, and well suited to the low frequency emulation of \ac{ASIC} based oscillators that are themselves linear in frequency. In sharing the \ac{FPGA} as a clock source again the pseudo-deterministic characteristics return, once more meaning this oscillator is better used for testing, simulating or verifying other blocks in the system.

\subsection{Ring Oscillator}
The best approximation of the non-idealities of \iac{ASIC} \ac{DCO} implementation can be obtained using \iac{RO}, also know as an inverter ring, as the \ac{DCO}. In \iac{ASIC} \iac{RO} is constructed by connecting a number of NOT gates back in a chain, with the output of the last NOT gate fed input the input of the first. As NOT gates are not available on \iac{FPGA}, they are replaced by inverter primitives. The inverter ring is an inherently unstable circuit that leverages the propagation delay through an odd number of inverters to create a waveform that, when viewed at a fixed position, is a square wave. 
\begin{figure}[h]
	\centering
	\includegraphics[width=0.4\textwidth]{../RO-wiki}
	\caption[Basic ring oscillator \ac{RTL} diagram]{Basic ring oscillator \ac{RTL} diagram \cite{ro_wiki}.}
	\label{fig:ro_wiki}
\end{figure}

@@

The main advantage of this type of oscillator stems from its similarity to that used on \iac{ASIC} which leads to many non-idealities carrying over from the truly mixed-signal implementation. Rather than being just a tool to validate that other modules, the performance of this oscillator can be analysed more deeply as well as the performance of networks containing it. Koskin \textit{et al} used such a design to test theories proposed regarding the synchronisation of oscillators during his PhD using this type of oscillator implemented on \iac{FPGA} \cite{koskin2019phd,theboys2019}. Also in this design's favour is the period step size, four times the inverter propagation delay, which was found experimentally to be than in the region of a nanosecond on Xilinx Artix-7 and Zynq \acp{FPGA}.

While it is advantageous for the purposes of analysis that this design is affected by variation in temperature, implementation and other non-idealities this brings with it challenges not present in a true mixed-signal design. In a design intended for \iac{ASIC} the design has near total control over the placement of circuit elements, however, the \ac{FPGA} based counterpart control is available over just the area in which the oscillator may lie, and the \ac{EDA} is responsible for the exact placement. Unlike the \ac{ASIC} implementation, in which the propagation delay through each element of the ring is well bounded, the lack of control over placement means that the propagation time through each inverter is not well bounded and may vary wildly. This makes it time consuming to ensure that the frequency tuning ranges of each oscillator match that desired, and use of this oscillator less straight-forward than that of the previous two designs. This is compounded by the simulation tools in \ac{EDA} packages intended for \acp{FPGA} being unable to accurately simulate the time delays, thus negating any benefit of test benches for the alignment process.

\section{Frequency Divider}
Many \aclp{PLL} (\acsp{PLL}), be they digital or analogue, incorporate a frequency divider in the feedback path so that comparison can be made with a reference operating at a lower frequency. This is a feature that is advantageous for \ac{ADPLL} networks, as comparison using this divided clock allows the synchronisation signal sent between modules to be of a frequency orders of magnitude lower than that of the clock distribution network and thus require circuitry that consumes significantly less energy. This division can be obtained trivially on \iac{FPGA} by a number of means, of which this thesis will mention just two. The first of these methods allows for integer division, by inverting a signal each time a counter reaches a certain value, $k$. When this condition is true the value stored in the counter is also reset to zero, thus implementing division of the generated clock frequency by $2k$.
\begin{figure}[h]%
	\centering
	\subfloat[Divider 1]{{\includegraphics[width=0.6\textwidth]{../divider1} }}\\
	\subfloat[Divider 2]{{\includegraphics[width=0.6\textwidth]{../divider2} }}%
	\caption[Frequency Divider \ac{RTL} Diagrams]{Frequency Divider \ac{RTL} Diagrams.}
	\label{fig:divs}
\end{figure}
By limiting to division by powers of two, the divider can be simplified with the removal of any comparison logic. One flip-flop and an inverter are required per power of two. As in Figure \ref{fig:divs} (b), the inverted $\overline{Q}$ output of each flip-flop is inverted and connected to the $D$ input, with the non-inverted $Q$ of the previous flip-flop acting as the clock. The first flip-flop in the chain is clocked by the \ac{DCO} output.

\section{Phase Detector}
Three potential \acl{PD} designs will be addressed in this thesis, sign-only detection with a Bang-Bang Detector, \ac{FPGA} clocked detection using \iac{FSM} and an emulation of \iac{TDL} using inverter primitives to create a delay, similarly to the \ac{RO} seen previously. Both designs will have to have a representation for both situations where the generated signal leads and lags the reference signal, and for the magnitude of this difference, a problem most obviously solved by using a signed binary representation of the difference. Most easily implemented in hardware is two's complement, as operations can be carried out by the same hardware as unsigned arithmetic. The sign convention for phase error assigns the positive values to the situation where the generated signal lags the reference, and vice versa for the leading case. Apart from the Bang-Bang Detector, each design features two blocks, one to determine the sign of the error and a second which estimates the magnitude. These blocks could theoretically be mixed-and-matched, but as they operate on different paradigms no benefit would be obtained.

\subsection{Bang-Bang Detector}
A Bang-Bang Detector is a sign-only detector, thus outputting a phase error just a single bit wide. This can be accomplished using a single flip-flop, with the reference signal acting as the clock, in this case using the rising edge, a design verified by Predraig in his 2011 thesis \cite{predraig}. The generated signal is applied to the $D$ input, thus when a rising edge is seen on the reference the value of the generated signal is held as the $Q$ output until the next rising edge of the reference. If the generated signal has gone high prior to the reference, and leading it, the value of $Q$ will be 1, thereby matching the expected sign behaviour. Trivially it can be seen that the case where the generated signal lags the reference the $Q$ will show 0.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.4\textwidth]{../BB}
	\caption[Bang-bang phase detector \ac{RTL} diagram]{Bang-bang phase detector \ac{RTL} diagram.}
	\label{fig:bang_bang}
\end{figure}

This design is simple but effective, and can even be used without a loop filter, albeit with a very limited oscillator tuning range. This is especially valuable an attribute at an early stage of platform development, as it may be used in order to test that other aspects of the system work as intended. The binary resolution of this detector, however, restricts its usefulness as the \ac{ADPLL} cannot react more quickly to larger phase differences.

\subsection{\acs{FPGA} Clocked Phase Detector}
The easier to implement of the two designs uses a combination of a state machine to determine which edge occurred first, and to control a two's complement up-down counter, which acts as the \ac{TDC} in this detector, and estimates the magnitude of the error.

\subsubsection{State Machine Based Detection}
While sign detection clocked by the \ac{FPGA} can be performed in a variety of ways, a state machine is a straightforward way in which to accomplish this. \Iacl{FSM} consists of three main components: the states themselves, the next state logic and the output logic. In this case the next state logic is the generated and reference signals, and the output logic sets the count instructions for the up-down counter as well as ensuring the phase error output is held constant between measurement periods. The \ac{FSM} will have a state representing the case where either the reference or generated signals occurred first, in addition to a wait state used between the end of one measurement interval and the commencement of the next. Two types of \ac{FSM} exist, Moore and Mealy Machines, the difference being that Moore Machines base their output only on the current state while Mealy Machines do so on both the current state and the inputs to the system. If a Moore Machines is used an extra state is required in order to hold the value of the phase error at the end of the measurement interval.

@@

There are three situations that may occur while the \ac{FSM} is in the waiting state. Firstly the case where the reference signal sees a rising edge before the generated clock, and thus the generated clock is said to be lagging the reference. As previously mentioned the convention dictates that the phase difference have a positive sign, so the counter is incremented to count in the positive direction. Similarly if the generated signal is leading the reference the sign of the error is negative and thus the counter is instructed to count downwards. The third potential scenario occurs when both clock signals have a rising edge within the same period of the \ac{FPGA} clock, which due to quantisation will be interpreted as both signals occurring at exactly the same time. In this case the counter receives no instruction and a zero value is stored until the next measurement interval. Once the \ac{FSM} is in a measurement state, and the counter either incrementing or decrementing, a rising edge occurring on the opposite signal to that which initiated the measurement will end the interval.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{../state_trans_new}
	\caption[Example State Transition Diagram for a Moore Machine]{Example State Transition Diagram for a Moore Machine.}
	\label{fig:state_trans}
\end{figure}

\subsubsection{Up-Down Counter}
The up-down counter in this design performs estimates the magnitude of the phase difference between the reference and generated signals. The state machine provides control instructions, indicating whether the count should be in the increasing or decreasing directions, and and holds the counter in reset once the measurement interval has been completed. The counter counts in two's complement, so counting in the downward direction can be carried out by adding on the representation of -1, the binary signal containing all 1s. The other advantage of counting in two's complement is that count value is already in the correct format for the follow block, the phase detector in a regular \ac{ADPLL} or the error combiner in this instance, therefore the count value at the end of the measurement interval can be stored in a register, which maintains the output value of the phase detector between measurements. Two's complement is an advantageous way to represent signed numbers in hardware as the mathematical operations are identical to their unsigned counterparts, thus allowing use of any adder or carry hardware that may be present on the \ac{FPGA}. An important attribute of this counter is the ability to freeze the count at the minimum and maximum count values in order to avoid a situation where the counter overflows, thus at high phase offsets potentially alternating between positive and negative differences which may restrict the locking range of the \ac{ADPLL}. Disallowing overflow also aligns the behaviour of this phase detector with that of those implementing delay lines, allowing better interchangeability and comparison, which for a testing platform may be valuable.

@@

\subsection{SigNum Detector}\label{section:signum}
The SigNum detector is based on the methodology used in the Bang-Bang \acl{PD} seen previously, but rather than just performing sign detection, the magnitude of the phase difference can also be measured. In the case of the \ac{FPGA} clocked design the counter handled the representation of both sign and magnitude, but as the name SigNum (Sign Number) suggested both are handled separately in this design. The modified Bang-Bang detector establishes which signal occurred first and generates a control signal that is analogous to the time delay between a rising edge on each input signal. \Iac{TDL} then uses this control signal to convert the phase difference between the signals to a digital signal. This magnitude is combined with the sign detected by the bang-Bang detector to form a two's complement error signal.

This method of phase detection is advantageous as similarly to the \acl{RO}, the \ac{TDL} can be implemented using inverters, and therefore is not synchronised to the \ac{FPGA} clock and the propagation time through inverters sets the resolution, thus allowing a smaller step than that of the clocked designs. The disadvantages, however, are also consistent with the \ac{RO} in that the propagation time will vary significantly depending on the \ac{EDA}'s placement of the circuit. This is not as severe a limitation for the \ac{TDL} as the \ac{RO} as while the tuning range is affected by this variation it cannot cause the centre point to vary, whereas the impact of placement on the \ac{RO} caused both variation in range and centre frequency.

>>>>

\section{Chapter Overview}
The first step in creating \iac{FPGA} based network of \acp{ADPLL} is the design of the \ac{ADPLL} itself, which will be addressed in this chapter. The nature of \iac{FPGA} necessitates compromises in the design of a certain blocks, limiting direct transfer from \ac{ASIC} designs. In this chapter the potential designs for each individual block, or module, investigated will be explained and the case for their selection in \iac{FPGA} based \ac{ADPLL} made. A number of blocks have implemented by purely digital circuitry and as such can be transferred in their entirety from \iac{FPGA} and vice versa. However, those that will be used to emulate mixed-signal circuitry, such as the \ac{DCO} and \ac{PFD}, will be examined in greater detail. \Iac{RTL} diagram of the \ac{ADPLL} topology that will be used in the following sections is shown in Figure \ref{fig:network_adpll}.
\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{../network_adpll}
	\caption{\ac{ADPLL} designed for use in a Cartesian grid network.}
	\label{fig:network_adpll}
\end{figure}

It is Important to note that despite a Cartesian grid leading to four neighbours, only two phase detectors are required per oscillator. Borrowing matrix indexing, and choosing the oscillator located at index $(i,j)$, $i,j \neq 1,N$ where $N$ is the number of oscillators in each direction, comparison need only be done with the signals generated by oscillators located at $(i-1,j)$ and $(i,j-1)$. The comparison with oscillators $(i+1,j)$ and $(i,j+1$) will be carried out in those \acp{ADPLL}, and the negation of the errors measured, used in the \ac{ADPLL} located at $(i,j)$.

\section{Digitally Controlled Oscillators}
The choices made in the design of the \ac{DCO} have the greatest impact on the effectiveness of the overall platform and which use cases the \ac{ADPLL} is suitable for, as all performance benchmarks are done using the waveform this block generates. This project will address three distinct designs of \acp{DCO} suitable for implementation on an FPGA, two derived from the clocks generated by the \acp{FPGA} own distribution network and one generated independently of this clock, using a chain of inverters. These are not the only ways in which an oscillator could be synthesised on \iac{FPGA}, however, other designs were deemed to be unsuitable for extensible and portable implementations.
%TODO delete if space needed
A prime example of this, is the use of Xilinx proprietary \texttt{IODELAY} blocks to create an oscillator, as detailed in Xilinx Application Note XAPP872 \cite{iodelay}. The key idea here is that the bulk of the period is made up by the propagation time through one of the \texttt{IODELAY} blocks, which can be set at implementation time. This is combined with a section of an inverter chain, and a multiplexer is used modify the length of this segment, the output of which is fed back into the \texttt{IODELAY} block. This method was discarded as the number of \texttt{IODELAY} blocks is very limited, so expanding to a larger network would be impossible, and as they are all located around the edge of the chip, not suited to the construction of a Cartesian grid.
%TODO delete if space needed

The main issue with the creation of \acp{DCO} on \iac{FPGA}, is the inability to create mixed-signal circuits, such as those used on \iac{ASIC}. Therefore the \ac{FPGA} based oscillator must emulate the behaviour of a mixed-signal circuit in some way.

\subsection{\acs{FPGA} Driven, Linear Period \acs{DCO}}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../osc1}
	\caption{\acs{FPGA} driven, linear period \acs{DCO} \ac{RTL} Diagram.}
	\label{fig:osc1}
\end{figure}
The first design of \ac{DCO} to be examined, is of the type used by Zianbetov in his \ac{ADPLL} network test bed and relies on a counter driven by the clock manager on the \ac{FPGA} \cite{zianbetov2013phd}. At each event on the \ac{FPGA} provided clock, a counter is incremented, overflowing upon increment past the maximum possible value. The \ac{MSB} of this counter forms the waveform generated by this oscillator, the period of which is controlled through an adjustable value that is loaded into the counter once overflow is reached, and forms the starting point for the counter. The period of oscillation is given by:
\begin{equation}
	T_{osc} = \big(2^{width} -~(BIAS+CC)~\big)\times T_{FPGA}
\end{equation}
where $T_{FPGA}$ is the clock period of the \ac{FPGA}, $CC$ is the control code input, $BIAS$ centres the oscillator in the middle of the tuning range in the event that the control code is zero and $width$ is the width of the counter used. As the only variable here is the control code, the period step of this design is $T_{FPGA}$, thereby providing period linearity with respect to the control code. This is the key advantage of this design, as most \ac{ASIC} implementations of \iac{DCO} are also linear in period. The other main reason to choose this design, is that its \ac{FPGA} clocked nature allows for exact control over the frequency of operation, and the number of tunable parameters make it possible to configure multiple ways to achieve the same frequencies of operation. Combined, these attributes make it very easy to create an oscillator that emulates the behaviour of a design intended for \iac{ASIC}, however, at a greatly reduced frequency. This restriction on the frequency of operation arises out of the period step size, which in order to obtain a good resolution must be orders of magnitude smaller than the intended period. As the output waveform is taken from the counter's \ac{MSB}, the reload value of the counter must never go beyond $2^{width-1}$, as otherwise the output waveform will become a constant 1. As the reload value varies the low time of the \ac{MSB}, if the desired output waveform is a square wave, this design will not be suitable.

Being \ac{FPGA} clocked this design has pseudo-deterministic characteristics, with each period step being almost identical across oscillators and over the entire tuning range, unlike \iac{ASIC} where process variation will impact the layout of a high frequency oscillator. The only variation in this design will come, ironically, from jitter or skew in the \ac{FPGA}'s clock distribution network, which as the frequencies will be in the hundreds of MHz is very minor. In the case of the Xilinx \acl*{Nexys} this is at most 100 picoseconds, or 0.05\% of the period of an intended output clock at 5 MHz. To put this value into perspective, on this board the minimum value of $T_{FPGA}$ that could be used to drive this oscillator is $3.875~\si{\nano\second}$, $1.935\%$ of the period.

@@

\subsection{\acs{FPGA} Driven, Linear Frequency \acs{DCO}}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../osc2}
	\caption{\acs{FPGA} driven, linear frequency \acs{DCO} \ac{RTL} Diagram.}
	\label{fig:osc2}
\end{figure}
The second \ac{FPGA} clocked oscillator is similar in most attributes to the above design but eschews period linearity for frequency linearity. Again the overflow property of a counter is used with the counter's \ac{MSB} as the block's output, however, this time it forms a square wave. Rather than setting the reload value of the counter, instead the increment is adjusted depending on the control code, thus requiring $\frac{2^N}{BIAS+CC}$ increments to overflow. Accordingly the frequency of operation is set by:
\begin{equation}
	f_{osc} = f_{FPGA}\times\frac{BIAS+CC}{2^{width}}
\end{equation}
Here the control code $CC$ and $BIAS$ are added to the value stored in the accumulator at each event of the \ac{FPGA} clock, until overflow is reached. This occurs at $2^{width}$ where, as before, $width$ is the bit width of the counter, thus valuing each control code increment at $\frac{f_{FPGA}}{2^{width}}$ Hz. As with the previous design, this oscillator is better suited to frequencies where the output of the \ac{DCO} is orders of magnitude lower than the clock signal driving it, as this ensures that the incremental change due to the control code remains a small fraction of the period.

This design is just as configurable as its linear-in-period counterpart, and well suited to the low frequency emulation of \ac{ASIC} based oscillators that are themselves linear in frequency. In sharing the \ac{FPGA} as a clock source, the pseudo-deterministic characteristics return, once more meaning this oscillator is better used for testing, simulating or verifying other blocks in the system.

\subsection{Ring Oscillator}
The best approximation of the non-idealities of \iac{ASIC} \ac{DCO} implementation can be obtained using \iac{RO}, also known as an inverter ring, as the \ac{DCO}. In \iac{ASIC} \iac{RO} is constructed by connecting a number of inverters back in a chain, with the output of the last NOT gate fed into the input of the first. As true inverters are not available on \iac{FPGA}, they are replaced by inverter primitives, constructed from look-up tables. The inverter ring is an inherently unstable circuit, that leverages the propagation delay through an odd number of inverters to create a waveform that, when viewed at a fixed position, is a square wave. 
\begin{figure}[h]
	\centering
	\includegraphics[width=0.4\textwidth]{../RO-wiki}
	\caption[Basic ring oscillator \ac{RTL} diagram]{Basic ring oscillator \ac{RTL} diagram \cite{ro_wiki}.}
	\label{fig:ro_wiki}
\end{figure}

@@

The main advantage of this type of oscillator stems from its similarity to that used on \iac{ASIC} which leads to many non-idealities carrying over from the truly mixed-signal implementation. Rather than being just a tool used to validate other modules, the performance of this oscillator can be analysed more deeply as well as the performance of networks containing it. Koskin \textit{et al} tested theories proposed regarding the synchronisation of oscillators during his PhD using this type of oscillator implemented on \iac{FPGA} \cite{koskin2019phd,theboys2019}. Also in this design's favour is the period step size, four times the inverter propagation delay, which was found experimentally to be within in the region of a nanosecond on Xilinx Artix-7 and Zynq \acp{FPGA}.

While it is advantageous for the purposes of analysis that this design is affected by variation in temperature, implementation and other non-idealities, this brings with it challenges not present in a true mixed-signal design. In a design intended for \iac{ASIC}, the designer has near total control over the placement of circuit elements, however, the \ac{FPGA} based counterpart control the designer may only set the region within the oscillator will lie, and the \acs{EDA} tool is responsible for the exact placement. Unlike the \ac{ASIC} implementation, in which the propagation delay through each element of the ring is well bounded, the lack of control over placement means that the propagation time through each inverter is not well bounded and may vary wildly. This makes it time consuming to ensure that the frequency tuning ranges of each oscillator match that desired, and making use of this oscillator less straight-forward than that of the previous two designs. This is compounded by the simulation tools in the \ac{EDA} packages intended for \acp{FPGA} being unable to accurately simulate the time delays, thus negating any benefit of test benches for the alignment process.

\section{Frequency Divider}
Many \aclp{PLL} (\acsp{PLL}), be they digital or analogue, incorporate a frequency divider in the feedback path, so that comparisons can be made with a reference operating at a lower frequency. This is a feature that is advantageous for \ac{ADPLL} networks, as comparisons using this divided clock allows the synchronisation signal sent between modules to be of a frequency orders of magnitude lower than that of the clock distribution network, and thus requiring circuitry that consumes significantly less energy. This division can be obtained trivially on \iac{FPGA} by a number of means; of which this thesis will mention just two. The first of these methods allows for integer division, by inverting a signal each time a counter reaches a certain value, $k$. When this condition is true the value stored in the counter is also reset to zero, thus dividing the generated clock frequency by $2k$.
\begin{figure}[h]%
	\centering
	\subfloat[Divider 1]{{\includegraphics[width=0.6\textwidth]{../divider1} }}\\
	\subfloat[Divider 2]{{\includegraphics[width=0.6\textwidth]{../divider2} }}%
	\caption[Frequency divider \ac{RTL} diagrams]{Frequency divider \ac{RTL} diagrams.}
	\label{fig:divs}
\end{figure}

By limiting division to powers of two, the second option ca simplify the divider by the removal of any comparison logic. One flip-flop and an inverter are required per power of two. As in Figure \ref{fig:divs} (b), the inverted $\overline{Q}$ output of each flip-flop is inverted and connected to the $D$ input, with the non-inverted $Q$ of the previous flip-flop acting as the clock. The first flip-flop in the chain is clocked by the \ac{DCO} output.

\section{Phase Detector}
Three potential \acl{PD} designs will be addressed in this thesis, sign-only detection with a Bang-Bang Detector, \ac{FPGA} clocked detection using \iac{FSM} and an emulation of \iac{TDL} using inverter primitives to create a delay, similarly to the \ac{RO} seen previously. All designs will have to have a representation for both situations where the generated signal leads and lags the reference signal, and for the magnitude of this difference. Both can be combined using a signed binary representation of the difference, such as two's complement. Two's complement is easily implemented, as operations can be carried out by the same hardware as unsigned arithmetic. The sign convention for phase error assigns the positive values to the situation where the generated signal lags the reference, and vice versa for the leading case. Apart from the Bang-Bang Detector, each design features two blocks, one to determine the sign of the error and a second to estimate the magnitude. These blocks could theoretically be mixed-and-matched, but as they operate on different paradigms, no benefit would be obtained.

\subsection{Bang-Bang Detector}
A Bang-Bang Detector is a sign-only detector, thus outputting a phase error a single bit wide. This can be accomplished using a single flip-flop, with the reference signal acting as the clock, in this case using the rising edge, a design verified by Predraig in his 2011 thesis \cite{predraig}. The generated signal is applied to the $D$ input, thus when a rising edge is seen on the reference, the value of the generated signal is held as the $Q$ output until the next rising edge of the reference. If the generated signal has gone high prior to the reference, and leading it, the value of $Q$ will be 1, thereby matching the expected sign behaviour. Trivially it can be seen that the case where the generated signal lags the reference the $Q$ will show $0$.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.4\textwidth]{../BB}
	\caption[Bang-bang phase detector \ac{RTL} diagram]{Bang-bang phase detector \ac{RTL} diagram.}
	\label{fig:bang_bang}
\end{figure}

This design is simple but effective, and can be used without a loop filter, albeit with a very limited oscillator tuning range. This is an especially valuable attribute at early stages of platform development, as it may be used in order to test that other aspects of the system work as intended. The binary resolution of this detector, however, restricts its usefulness ,as the \ac{ADPLL} cannot react more quickly to larger phase differences.

\subsection{\acs{FPGA} Clocked Phase Detector}
The easier to implement of the two designs, uses a combination of a state machine to determine which edge occurred first, and to control a two's complement up-down counter, which acts as the \ac{TDC} in this detector, and estimates the magnitude of the error.

\subsubsection{State Machine Based Detection}
While sign detection clocked by the \ac{FPGA} can be performed in a variety of ways, a state machine is a straightforward way in which to accomplish this. \Iacl{FSM} consists of three main components: the states themselves, the next state logic and the output logic. In this case the next state logic is the generated and reference signals, and the output logic sets the count instructions for the up-down counter, as well as ensuring the phase error output is held constant between measurement periods. The \ac{FSM} will have a state representing the case where either the reference or generated signals occurred first, in addition to a wait state used between the end of one measurement interval and the commencement of the next. Two types of \ac{FSM} exist, Moore and Mealy Machines, the difference being that Moore Machines base their output only on the current state, while Mealy Machines do so on both the current state and the inputs to the system. If a Moore Machines is used an extra state is required in order to hold the value of the phase error at the end of the measurement interval.

@@

There are three situations that may occur while the \ac{FSM} is in the waiting state. Firstly, the case where the reference signal sees a rising edge before the generated clock, and thus the generated clock is said to be lagging the reference. As previously mentioned, the convention dictates that the phase difference have a positive sign, so the counter is incremented to count in the positive direction. Secondly, if the generated signal is leading the reference, the sign of the error is negative and thus the counter is instructed to count downwards. The third potential scenario occurs when both clock signals have a rising edge within the same period of the \ac{FPGA} clock, which due to quantisation will be interpreted as both signals occurring at exactly the same time. In this case the counter receives no instruction and a zero value is stored until the next measurement interval. Once the \ac{FSM} is in a measurement state, and the counter either incrementing or decrementing, a rising edge occurring on the opposite signal to that which initiated the measurement, will end the interval.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{../state_trans_new}
	\caption[Example State Transition Diagram for a Moore Machine]{Example State Transition Diagram for a Moore Machine.}
	\label{fig:state_trans}
\end{figure}

\subsubsection{Up-Down Counter}
The up-down counter in this design estimates the magnitude of the phase difference between the reference and generated signals. The state machine provides control instructions, indicating whether the count should be in the increasing or decreasing directions, and holds the counter in reset once the measurement interval has been completed. The counter counts in two's complement, so counting in the downward direction can be carried out by adding on the representation of -1, the binary signal containing all 1s. The other advantage of counting in two's complement is that count value is already in the correct format for the follow block, the phase detector in a regular \ac{ADPLL} or the error combiner in this instance, therefore the count value at the end of the measurement interval can be stored in a register, which maintains the output value of the phase detector between measurements. Two's complement is an advantageous way to represent signed numbers in hardware as the mathematical operations are identical to their unsigned counterparts, thus allowing use of any adder or carry hardware that may be present on the \ac{FPGA}. An important attribute of this counter is the ability to freeze the count at the minimum and maximum count values, in order to avoid a situation where the counter overflows. At high phase offsets this may cause alternation between positive and negative differences, restricting the locking range of the \ac{ADPLL}. Disallowing overflow also aligns the behaviour of this phase detector with that of those implementing delay lines, allowing better interchangeability and comparison, which for a testing platform is valuable.

@@

\subsection{SigNum Detector}\label{section:signum}
The SigNum detector is based on the methodology used in the Bang-Bang \acl{PD} seen previously, but rather than just performing sign detection, the magnitude of the phase difference can also be measured. In the case of the \ac{FPGA} clocked design, the counter handled the representation of both sign and magnitude, but as the name SigNum (Sign Number) suggests, both are handled separately in this design. The modified Bang-Bang detector establishes which signal occurred first and generates a control signal that is analogous to the time delay between a rising edge on each input signal. \Iacf{TDL} then uses this control signal to convert the phase difference between the signals to a digital signal. The magnitude is combined with the sign detected by the bang-Bang detector to form a two's complement error signal.

This method of phase detection is advantageous, as similarly to the \acl{RO}, the \ac{TDL} can be implemented using inverters, and therefore is not synchronised to the \ac{FPGA} clock. The propagation time through inverters once more sets the resolution, thus allowing a smaller step than that of the clocked designs. The disadvantages, however, are also consistent with the \ac{RO} in that the propagation time will vary significantly depending on the \ac{EDA}'s placement of the circuit. This is not as severe a limitation for the \ac{TDL} as the \ac{RO} as, while the tuning range is affected by this variation, it cannot cause the centre point to vary, whereas the impact of placement on the \ac{RO} caused both variation in range and centre frequency.
